---
title: "KMeans"
author: "Shafquat Arefeen"
date: "July 20, 2018"
output: html_document
---
## Loading the Data

The first step is to load the joined data and profile the data

```{r}
# load by browsing
seg.raw <- read.csv(file.choose())

# check column names 
names(seg.raw)

# Assign the last date of the data for age calculations
data_last_date <- as.Date("2017-12-31")
```

## Field Manipulation

Create new quantitative fields to run K-Means.
These fields are Male, Age, Account Age, and Last Activity
Dormant customers are removed as required. If the Last Activity is <= 1 then the customer is considered to be active as per the requirements.

```{r cars}
# Male Dummy
seg.raw$Male[seg.raw$gender=='Female'] <- 0
seg.raw$Male[seg.raw$gender=='Male'] <- 1

# Remove all non int entries in Male field
seg.clean <- seg.raw[!is.na(seg.raw$Male),]

# Add age field from birthdate
library('eeptools')
seg.clean$age <- floor(age_calc(as.Date(seg.clean$Birthdate), enddate = data_last_date, units = "years"))

# Remove Dormant members
seg.clean <- seg.clean[seg.clean$ActivityStatusScotiaScene != 2,]

# Add new field based on BlackEarnLastDt if year > 1 then inactive
seg.clean$last_activity <- age_calc(as.Date(seg.clean$BlackEarnLastDt), enddate = data_last_date, units = "years")

# Add account age field
seg.clean$account_age <- floor(age_calc(as.Date(as.character(seg.clean$AccountOpenKey), "%Y%m%d"), enddate = data_last_date, units = "months"))

```

## Active members only

Keep only the active members in the dataframe by using the last_activity field

```{r}
# Keep Active members
seg.active <- seg.clean[seg.clean$last_activity <= 1,]
```


## Create a function to Aggregate the fields by average

A simple function to report means by group

```{r}
seg.summ <- function(data, groups) {
  aggregate(data, list(groups), function(x) mean(as.numeric(x)))  
}
```


## Subset the fields for K-Means

Only select the quantitative fields for K-Means.

```{r}
# convert factor variables to numeric (kmeans requires). OK b/c all are binary.
seg.df.num <- seg.active[c(
                       'account_age',
                       'age',
                       'OnlineTicketPurchaser_value',
                       'TuesdayAttendee_value',
                       'ConcessionPurchaser_value',
                       'AttendsWithChild_value',
                       'WeekendMatineeViewer_value',
                       'WeekdayMatineeViewer_value',
                       'BlackEarnCount',
                       'BlackEarnPointTotal',
                       'BlackBurnCount',
                       'BlackBurnPointTotal',
                       'BlackActivityDays',
                       'VISACineplex',
                       'VISA',
                       'CineplexIssue',
                       'CineplexRedeem',
                       'PointsScene',
                       'CARAIssue',
                       'CARARedeem',
                       'Male',
                       'last_activity'
                       )]
summary(seg.df.num)
```

## Determine number of clusters

Scale the numbers so no fields are overweighted.
The goal is to minimize the within cluster variance. And plot in an elbow graph to decide how many clusters to choose.

```{r}
# Scale values
scaled_df <- scale(seg.df.num)

wss <- (nrow(scaled_df)-1)*sum(apply(scaled_df,2,var))
for (i in 2:15) wss[i] <- sum(kmeans(scaled_df, 
                                     centers=i)$withinss)

# The elbow graph to gauge the number of clusters
plot(1:15, wss, type="b", xlab="Number of Clusters",
     ylab="Within groups sum of squares")
```


## K-Means

Run K-Means and inspect the result.

```{r}
set.seed(96743) # set seed to make sure the random assignment starts at the same point
seg.k5 <- kmeans(scaled_df, centers=5)

# inspect the 4-segment solution by looking at the means per segment per numeric
seg.summ(seg.df.num, seg.k5$cluster)
```

## Output results
Create a csv for the results that can be used to get summary statistics.

```{r}
# assign Cluster to DataFrame
seg.df.num$cluster <- seg.k5$cluster

# assign user_id back to to DataFrame
seg.df.num$Unique_member_identifier <- seg.active$Unique_member_identifier
seg.df.num$age_class <- seg.active$age_class

# write dataframe to csv for summary statistics
write.csv(seg.df.num, file = "K_means_data.csv")
```



## Cluster Visualization
Run K-Means and inspect the result.

```{r}
library("flexclust")

set.seed(1)
dat <- seg.df.num
ind <- sample(nrow(dat), 1000)

dat[["train"]] <- TRUE
dat[["train"]][ind] <- FALSE

cl1 = kcca(dat[dat[["train"]]==TRUE, 1:2], k=5, kccaFamily("kmeans"))
cl1    

pred_train <- predict(cl1)
pred_test <- predict(cl1, newdata=dat[dat[["train"]]==FALSE, 1:2])

image(cl1)
points(dat[dat[["train"]]==TRUE, 1:2], col=pred_train, pch=19, cex=0.3)
points(dat[dat[["train"]]==FALSE, 1:2], col=pred_test, pch=22, bg="orange")
```

## Summary Segments
Get the Size of each cluster, gender by cluster, age_class by cluster, tenure enrollment source by cluster, activity at CPX in last year, activity at other partners in last year

```{r}
# Choose the exported dataset K_means_data.csv
sum.segments <- read.csv(file.choose())
```

Size of each cluster
```{r}
aggregate(sum.segments$Unique_member_identifier, by=list(sum.segments$cluster), FUN=length)
```
Gender by cluster
```{r}
aggregate(sum.segments$Male, by=list(sum.segments$cluster), FUN=sum)
```
Age_Class by cluster
```{r}
aggregate(sum.segments$Unique_member_identifier, by=list(sum.segments$cluster,sum.segments$age_class), FUN=length)
```
tenure enrollment source by cluster
```{r}
aggregate(sum.segments$Unique_member_identifier, by=list(sum.segments$cluster,sum.segments$account_age), FUN=length)
```
activity at CPX in last year
```{r}
round(aggregate(sum.segments$CineplexIssue, by=list(sum.segments$cluster), FUN=mean))
round(aggregate(sum.segments$CineplexRedeem, by=list(sum.segments$cluster), FUN=mean))
```
activity at other partners in last year
```{r}
round(aggregate(sum.segments$CARAIssue, by=list(sum.segments$cluster), FUN=mean))
round(aggregate(sum.segments$CARARedeem, by=list(sum.segments$cluster), FUN=mean))
```